---
title: "Day 3 - Morning"
author: "Alex Genovese"
date: "6/3/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rstan)
library(HDInterval)
library(scatterplot3d)
library(brms)
library(rstanarm)
library(scales)
```

```{r Setup, include=FALSE}
# Some constants ++++++++++++++++++++++++
n_chains <- 4
n_iter <- 5000
n_warmup <- 200
n_thin <- 1
# Only if you have a mutlicore machine
# Only worth it for complex models, otherwise slower
n_cores <- n_chains
# Read in the data ============================
d <- read.csv('Admission_Predict_Ver1.1.csv')
(N <- nrow(d))

# Graphics stuff ++++++++++++++++++++++++
par(mar=c(5.1, 5.1, 4.1, 3.1))
```
# Regression with a constant predictor

## Constant Predictor Model

* $y_{i} = \mu + \epsilon_{i}$ (y=mean+noise) 
$\epsilon_{i} \sim N(0,\sigma)$ (noise is a normal distribution with 0 mean and sigma standard deviation)
* **$y_{i} \sim N(\mu,\sigma)$** (y is a normal distribution with a $\mu$ mean and $\sigma$ standard deviation)

```{r Regression with a constant predictor, eval=FALSE, include=FALSE}

# Standardize the continuous variables ========
# Not strictly needed.
# In fact, might be counterindicated.
# Helps keep coefficients in a small, known range.
# Done here for pedagogical reasons.
# If not done, set initial parameters to ensure convergence.
d$TOEFL.Score_Stand <- as.numeric(scale(d$TOEFL.Score))
d$University.Rating_Stand <- as.numeric(scale(d$University.Rating))
d$SOP_Stand <- as.numeric(scale(d$SOP))
d$LOR_Stand <- as.numeric(scale(d$LOR)) 
d$CGPA_Stand <- as.numeric(scale(d$CGPA)) 
# How to untransform coefficients
# https://stats.stackexchange.com/questions/30876/how-to-convert-standardized-coefficients-to-unstandardized-coefficients

# I'm also standardizing the DV, for simplicity
# Some statisticians might cringe
d$GRE.Score_Stand <- as.numeric(scale(d$GRE.Score)) 

par(mfrow=c(2,1))
#hist(d$GRE.Score, xlab='Raw GRE', main='')
#hist(d$GRE.Score_Stand, xlab='Standardized GRE', main='')
par(mfrow=c(1,1))

# Explore the data =================================
head(d)
str(d)
summary(d)

panel.hist <- function(x, ...)
{
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(usr[1:2], 0, 1.5) )
  h <- hist(x, plot = FALSE)
  breaks <- h$breaks; nB <- length(breaks)
  y <- h$counts; y <- y/max(y)
  rect(breaks[-nB], 0, breaks[-1], y, col = "cyan", ...)
}
pairs(d[,c(8, 10:15)], lower.panel=NULL, diag.panel=panel.hist)

# =================================================
# Constant predictor (const) 
# =================================================

# Handle the data

data_list_const <- list( N = N, y = d$GRE.Score_Stand ) # The input to Stan
data_list_const

# Compile code to create dynamic shared object (DSO)

# stan_dso_const <- stan_model( file = 'regression_const.stan' )
# save('stan_dso_const', file = 'stan_dso_const.RData') # Save it to load faster later
load(file = 'stan_dso_const.RData') # Load a saved model

# ==========================================
# Generate Monte Carlo samples 

# Initial params
init_params_const <- 'random' # The default
# init_params_const <- list(n_chains)
# for (i in 1:n_chains) {
#   init_params_const[[i]] <- list( mu = 316,
#                                   sigma = 10 )
# }
# init_params_const

stan_fit_const <- sampling( object = stan_dso_const,
                            data = data_list_const,
                            chains = n_chains,
                            iter = n_iter,
                            warmup = n_warmup,
                            thin = n_thin,
                            cores = n_cores,
                            init = init_params_const )

# Diagnostics ---------------------

# Chain mixing
stan_trace(stan_fit_const, pars = c('mu', 'sigma'))

# Autocorrelation
stan_ac(stan_fit_const, pars = c('mu', 'sigma'))

# Effective sample size
# Ratio of effective sample size to total sample size.
# You want this to be close to 1, definitely > .10.
stan_ess(stan_fit_const, pars = c('mu', 'sigma'))

# Monte Carlo Standard Error
# The standard error of the mean of the posterior draws. 
# Want mcse than 10% of the posterior standard deviation.
stan_mcse(stan_fit_const, pars = c('mu', 'sigma'))

# Rhat statistic
# The ratio of the average variance of samples within each chain to the 
#   variance of the pooled samples across chains; if all chains are at equilibrium, 
#   these will be the same and R̂ will be one.
# You want Rhat to be less than 1 (recommended < 1.05)
stan_rhat(stan_fit_const, pars = c('mu', 'sigma'))

# Kernel density estimates
stan_dens(stan_fit_const, pars = c('mu', 'sigma'), separate_chains = TRUE, alpha = 0.3)

# =================================================
# Posteriors

# Numeric output 
print(stan_fit_const, pars = c('mu', 'sigma') )
mean(d$GRE.Score_Stand)
sd(d$GRE.Score_Stand)
summary(stan_fit_const, pars = c('mu', 'sigma') )[[1]]

# Histgrams
# Priors & Posteriors
stan_plot(stan_fit_const, pars = c('mu_prior', 'mu'), point_est = "mean", show_density = TRUE, fill_color = "maroon")
stan_plot(stan_fit_const, pars = c('sigma_prior', 'sigma'), point_est = "mean", show_density = TRUE, fill_color = "maroon")

stan_hist(stan_fit_const, pars = c('mu_prior', 'mu'), bins = 30)
stan_hist(stan_fit_const, pars = c('sigma_prior', 'sigma'), bins = 30)

# Posterior predictive check 

# Extract the mcmc samples ---------------------

# y's
# A complete set of 500 for each mu and sigma
y_post_const <- extract(stan_fit_const, 'y_pred')[[1]]

# residuals
# A complete set of 500 for each mu and sigma
resid_post_const <- extract(stan_fit_const, 'resid')[[1]]

# Generate and graph posterior ---------------------

# y's for 5 representative samples
par(mfrow=c(3, 2))

#hist(d$GRE.Score_Stand, xlab='gre', 
     cex.lab=1.25, main='', 
     prob=TRUE, 
     xlim=c(-3, 3), 
     col='green')
legend(x='topleft', legend=c('y', 'y_pred'), cex=1.5, col=c('green', 'gray'), pch=c(15, 15), lty=c(0,0), bty='n')

for (i in 1:5) {
  hist(y_post_const[i,], xlab='gre', 
       cex.lab=1.25, main='', 
       prob=TRUE, 
       xlim=c(-3, 3), 
       col='gray')
}

par(mfrow=c(1, 1))

# (Boring) Residual plot ---------------------

# residuals for 5 representative samples
par(mfrow=c(3, 2))

plot(d$GRE.Score_Stand - mean(d$GRE.Score_Stand),
     xlab='index', ylab='residual',
     col='green', cex.lab=1.5, cex.axis=1.5, pch=16)
abline(h=0)

for (i in 1:5) {
  plot(resid_post_const[i,],
       xlab='index', ylab='residual',
       col='gray', cex.lab=1.5, cex.axis=1.5, pch=16)
  abline(h=0)
}

par(mfrow=c(1, 1))

```

# Intro to rstanarm

## rstanarm

* rstan+applied regression modeling
* Develped by the stan team
* Good for basic to intermediate models
* Pre-compiled Stan code
  + Runs very quickly
  
```{r rstanarm, eval=FALSE, include=FALSE}
# ======================================
# rstanarm
# ?rstanarm

# Reminder of the standardized GRE scores
hist(d$GRE.Score_Stand, xlab='Standardized GRE', main='')

# First, with standardized DV for comparison to Stan
# ~ here mean 'as a function of'
rstanarm_const_stand <- stan_glm(GRE.Score_Stand ~ 1,
                                 data=d,
                                 family = gaussian())
summary(rstanarm_const_stand, digits=3, probs=c(.025, .5, .975)) # Summary
print(stan_fit_const, pars = c('mu', 'sigma') ) # From using STAN

# Now, with unstandardized DV
rstanarm_const_raw <- stan_glm(GRE.Score ~ 1,
                               data=d,
                               family = gaussian())
summary(rstanarm_const_raw, digits=3, probs=c(.025, .5, .975)) # Summary

```

## brms
* Bayesian regression model
* Developed in collaboration of stan
* Good for basic to complex models
* Not pre-compiled
  + Slower
  + More flexible


# Bivariate regression
## One predictor

* $y_{i} = \beta_{0}+\beta_{1}x_{i}+\epsilon_{i}$
  + $\epsilon \sim N(0,\sigma)$
  
* $\mu_{i}=\beta_{0}+\beta_{1}x_{i}$

* **$y_{i} \sim N(\beta_{0}+\beta_{1}x_{i}, \sigma)$**

```{r eval=FALSE, include=FALSE}
# =================================================
# Single predictor (1_pred) 
# =================================================

# Handle the data 

N_subset <- N # N = all data
data_subset <- sample(1:N, N_subset) # Take a subset, to reduce N, if desired
data_list_1_pred <- list( N = N_subset, 
                          y = d$GRE.Score_Stand[data_subset], 
                          x1 = d$CGPA_Stand[data_subset] ) # The input to Stan

# Compile code to create dynamic shared object (DSO) ++++++++++++++++++++++++
#stan_dso_1_pred <- stan_model( file = 'regression_1_pred.stan')
#save('stan_dso_1_pred', file = 'stan_dso_1_pred.RData') # Save the model
load('stan_dso_1_pred.RData') # Load a saved model

# Generate Monte Carlo samples ++++++++++++++++++++++++

# Initial params
init_params_const <- 'random' # The default
# init_params_const <- list(n_chains)
# for (i in 1:n_chains) {
#   init_params_const[[i]] <- list( b0 = 100,
#                                   b1 = 200,
#                                   sigma = 10 )
# }
# init_params_const

stan_fit_1_pred <- sampling( object = stan_dso_1_pred,
                             data = data_list_1_pred,
                             chains = n_chains,
                             iter = n_iter,
                             warmup = n_warmup,
                             thin = n_thin,
                             cores = n_cores,
                             init = init_params_const)

# Diagnostics 
# Chain mixing
stan_trace(stan_fit_1_pred, pars = c('b0', 'b1', 'sigma'))

# Autocorrelation
stan_ac(stan_fit_1_pred, pars = c('b0', 'b1', 'sigma'))

# Effective sample size
# Ratio of effective sample size to total sample size.
# You want this to be close to 1, definitely > .10.
stan_ess(stan_fit_1_pred, pars = c('b0', 'b1', 'sigma'))

# Monte Carlo Standard Error
# The standard error of the mean of the posterior draws. 
# Want mcse than 10% of the posterior standard deviation.
stan_mcse(stan_fit_1_pred, pars = c('b0', 'b1', 'sigma'))

# Rhat statistic
# The ratio of the average variance of samples within each chain to the 
#   variance of the pooled samples across chains; if all chains are at equilibrium, 
#   these will be the same and R̂ will be one.
# You want Rhat to be less than 1 (recommended < 1.05)
stan_rhat(stan_fit_1_pred, pars = c('b0', 'b1', 'sigma'))

# Kernel density estimates
stan_dens(stan_fit_1_pred, pars = c('b0', 'b1', 'sigma'), separate_chains = TRUE, alpha = 0.3)

# Posteriors ---------------------

# Numeric output 
print(stan_fit_1_pred, pars = c('b0', 'b1', 'sigma') )
summary(stan_fit_1_pred, pars = c('b0', 'b1', 'sigma') )[[1]]

# Histograms
stan_plot(stan_fit_1_pred, pars = c('b0', 'b1', 'sigma'), point_est = "mean", show_density = TRUE, fill_color = "maroon")

# Compare to OLS regression
lm_1_pred <- lm(GRE.Score_Stand[data_subset] ~ CGPA_Stand[data_subset], data=d)
summary(lm_1_pred)
confint(lm_1_pred)
print(stan_fit_1_pred, pars = c('b0', 'b1', 'sigma') )

# Histgrams
# Priors & Posteriors
stan_plot(stan_fit_1_pred, pars = c('b0_prior', 'b0'), point_est = "mean", show_density = TRUE, fill_color = "maroon")
stan_plot(stan_fit_1_pred, pars = c('b1_prior', 'b1'), point_est = "mean", show_density = TRUE, fill_color = "maroon")
stan_plot(stan_fit_1_pred, pars = c('sigma_prior', 'sigma'), point_est = "mean", show_density = TRUE, fill_color = "maroon")

# Scatterplot of samples
# Priors
pairs(stan_fit_1_pred, pars = c('b0_prior', 'b1_prior', 'sigma_prior'))
# Posteriors
pairs(stan_fit_1_pred, pars = c('b0', 'b1', 'sigma'))

# Posterior predictive check ++++++++++++++++++++++++

# Extract the mcmc samples ---------------------
x_post_1_pred <- extract(stan_fit_1_pred, 'x_pred')[[1]]
y_post_1_pred <- extract(stan_fit_1_pred, 'y_pred')[[1]]

# y's for 5 representative samples
par(mfrow=c(3, 2), pty='s')

plot(d$CGPA_Stand, d$GRE.Score_Stand, 
     xlab='cgpa', ylab='gre', 
     cex.lab=1.25, main='', 
     xlim=c(-3, 3), 
     col='green')
legend(x='topleft', legend=c('y', 'y_pred'), cex=1.5, col=c('green', 'gray'), pch=c(16, 16), lty=c(0,0), bty='n')

for (i in 1:5) {
  plot(x_post_1_pred[i,], y_post_1_pred[i,], 
       xlab='cgpa', ylab='gre', 
       cex.lab=1.25, main='', 
       xlim=c(-3, 3), 
       col='gray')
}

par(mfrow=c(1, 1), pty='m')

# Extract the mcmc samples ---------------------
mu_post_1_pred <- as.numeric(extract(stan_fit_1_pred, 'mu')[[1]])
b0_post_1_pred <- as.numeric(extract(stan_fit_1_pred, 'b0')[[1]])
b1_post_1_pred <- as.numeric(extract(stan_fit_1_pred, 'b1')[[1]])
sigma_post_1_pred <- as.numeric(extract(stan_fit_1_pred, 'sigma')[[1]])

# How many mc samples are there
(n_1_pred <- length(b0_post_1_pred))

# Variability in regression prediction ---------------------
plot(d$CGPA_Stand[data_subset], d$GRE.Score_Stand[data_subset],
     xlab='college GPA', ylab='GRE',
     cex.lab=1.4)

# Median regression line
# Note: Parameters are evaluated independently
abline(a=median(b0_post_1_pred), b=median(b1_post_1_pred), lwd=5, col='blue')

# Distributions of lines
# 20 samples from posterior
for (i in 1:20) {
  abline(a=b0_post_1_pred[i], b=b1_post_1_pred[i], col=alpha("blue", .3))
}

# Credible intervals ---------------------
plot(d$CGPA_Stand[data_subset], d$GRE.Score_Stand[data_subset],
     xlab='college GPA', ylab='GRE',
     cex.lab=1.4)

# Median regression line
abline(a=median(b0_post_1_pred), b=median(b1_post_1_pred), lwd=5, col='blue')

# For the mean

# mu_link determines the predicted mean mu for a given cgpa
# We could also do this within Stan (see generated quantities)
mu_link_1_pred <- function(cgpa) b0_post_1_pred + b1_post_1_pred*cgpa 
mu_link_1_pred(-1.5)[1:10]
b0_post_1_pred[1] + b1_post_1_pred[1]*-1.5

# The range of cpga to consider
cgpa_seq <- seq(from = -5, to = 5, by = .1)
cgpa_seq[1:10]

# Sample of mu for each cgpa
mu_seq_1_pred <- sapply(cgpa_seq, mu_link_1_pred)
mu_seq_1_pred[1:3, 1:10] # row=mcmc sample, col=mu sample for each cgpa

# Mean of mu for each cgpa
mu_mean_1_pred <- apply(mu_seq_1_pred, 2, mean)
mu_mean_1_pred[1:10] # Mean of all of the samples for that cgpa
mean(mu_seq_1_pred[,1])

# Highest posterior density interval for mu for each cgpa
mu_hpdi_1_pred <- apply(mu_seq_1_pred, 2, hdi, prob=.95)
hdi(mu_seq_1_pred[,36], prob=.95) # At cgpa = -1.5

# Draws the shaded region
# 95% HDI
polygon(c(cgpa_seq, rev(cgpa_seq)), 
        c(mu_hpdi_1_pred[2,], rev(mu_hpdi_1_pred[1,])), 
        col=alpha('blue', .3))

# For new gre values

# new_gre_link_1_pred samples a set of new values for a given cgpa
new_gre_link_1_pred <- function(cgpa) {
  rnorm(
    n = n_1_pred,
    mean = b0_post_1_pred + b1_post_1_pred*cgpa,
    sd = sigma_post_1_pred
  )
}
new_gre_link_1_pred(-1.5)[1:10]

# Calls new_value_link for each cgpa
new_gre <- sapply(cgpa_seq, new_gre_link_1_pred)
new_gre[1:3, 1:10] # row=mcmc sample, col=new value sample for each cgpa

# Highest posterior density interval for the new values for each cgpa
new_gre_hdpi <- apply(new_gre, 2, hdi, prob=.95)
hdi(new_gre_hdpi[,36], prob=.95) # At cgpa = -1.5

# Draws the shaded region
# 95% HDI
polygon(c(cgpa_seq, rev(cgpa_seq)), 
        c(new_gre_hdpi[2,], rev(new_gre_hdpi[1,])), 
        col=alpha('blue', .15))

# A legend
legend(x='bottomright',
       legend=c('data', 'median mu', 'HDPI for mean gre', 'HDPI for new gre'),
       pch=c(1, 15, 15, 15), 
       col=c('black', 'blue', alpha('blue', .3), alpha('blue', .15)),
       bty='n',
       cex=1.5)

# rstanarm ++++++++++++++++++++++++

# Unstandardized
rstanarm_1_pred <- stan_glm(GRE.Score ~ CGPA,
                            data=d)
summary(rstanarm_1_pred, digits=3, probs=c(.025, .5, .975)) # Summary

# Fit Diagnostics 
#   mean_PPD
#   The sample average posterior predictive distribution of the outcome. 
#   This is useful as a quick diagnostic.
#   A useful heuristic is to check if mean_PPD is plausible when compared to mean(y).
mean(d$GRE.Score)
# If it is plausible then this does not mean that the model is good,
#   only that it can reproduce the sample mean, 
#   however if mean_PPD is implausible then it is a sign that something is wrong.

# Other
#   mean_PPD: mean of the posterior predictive distribution (hopefully on par 
#     with the mean of the target variable).
#   log-posterior: similar to the log-likelihood from maximum likelihood, but 
#     for the Bayesian case.

# Diagnostics ----------------

print(rstanarm_1_pred)
posterior_interval(rstanarm_1_pred, prob = 0.95) # Posterior intervals
plot(rstanarm_1_pred, pars='(Intercept)') # On different plots b/c scales are so different
plot(rstanarm_1_pred, pars='sigma')

plot(rstanarm_1_pred, "trace") # Chains
plot(rstanarm_1_pred, "dens") # Posterior densities
plot(rstanarm_1_pred, "pairs") # Scatterplots
plot(rstanarm_1_pred, "scatter", pars=c('(Intercept)', 'CGPA')) # Only used for exactly 2 parameters
plot(rstanarm_1_pred, "neff") # neff
plot(rstanarm_1_pred, "rhat") # rhat
plot(rstanarm_1_pred, "acf") # Autocorrelation
plot(rstanarm_1_pred, "areas", prob = 0.95, prob_outer = 1)
plot(rstanarm_1_pred, "areas", pars=c('CGPA'), prob = 0.95, prob_outer = 1)

summary(residuals(rstanarm_1_pred)) # Residuals

# Posterior predictive check ----------------

posterior_predict(rstanarm_1_pred) # Posterior samples
posteriorSamples <- as.matrix(rstanarm_1_pred, pars='CGPA')
head(posteriorSamples)
hist(posteriorSamples)
mean(posteriorSamples > 15) # P(CGPA > 15)

pp_check(rstanarm_1_pred, nreps = 100) 
pp_check(rstanarm_1_pred, plotfun = "boxplot", nreps = 100, notch = FALSE)
pp_check(rstanarm_1_pred, plotfun = "hist", nreps = 6)
pp_check(rstanarm_1_pred, plotfun = "stat_2d", stat = c("mean", "sd"))
pp_check(rstanarm_1_pred, plotfun = "scatter_avg") 
pp_check(rstanarm_1_pred, plotfun = "scatter", nreps = 6)
pp_check(rstanarm_1_pred, plotfun = "intervals")
pp_check(rstanarm_1_pred, plotfun = "error_hist", nreps = 6)

# Shinystan!!! ----------------
# All-in-one COOL!

launch_shinystan(rstanarm_1_pred) 

# Priors ----------------

prior_summary(rstanarm_1_pred)
# Automatically mean centers predictors!
# Adjusted scale values are the prior scales actually used by rstanarm
#   and are computed by adjusting the prior scales specified by the user 
#   to account for the scales of the predictors 

# Intercept = N(0, 10*sy)
10*sd(d$GRE.Score)
# Coefficents = N(0, 2.5*sy/sx)
2.5 * sd(d$GRE.Score)/sd(d$CGPA)
# Sigma = exp(1/sy)
1/sd(d$GRE.Score)

# New HORRIBLE priors (to show how things go wrong)
my_coef_prior <- normal(location = c(-100), 
                        scale = c(.01), 
                        autoscale = FALSE)
my_int_prior <- normal(location = c(100), 
                       scale = c(.01), 
                       autoscale = FALSE)
my_aux_prior <- exponential(10,
                            autoscale = FALSE)
rstanarm_1_pred_prior <- stan_glm(GRE.Score ~ CGPA,
                                  data=d,
                                  family = gaussian(),
                                  prior = my_coef_prior,
                                  prior_intercept = my_int_prior,
                                  prior_aux = my_aux_prior)
summary(rstanarm_1_pred_prior, digits = 3)

prior_summary(rstanarm_1_pred_prior)

# Diagnostics look ok
plot(rstanarm_1_pred_prior, "trace") # Chains 
plot(rstanarm_1_pred_prior, "neff") # neff
plot(rstanarm_1_pred_prior, "rhat") # rhat
plot(rstanarm_1_pred_prior, "acf") # Autocorrelation
plot(rstanarm_1_pred_prior, "areas", prob = 0.95, prob_outer = 1)

# Posterior predictive looks not ok
pp_check(rstanarm_1_pred_prior, nreps = 10) 
pp_check(rstanarm_1_pred_prior, plotfun = "boxplot", nreps = 10, notch = FALSE) # Not so good
pp_check(rstanarm_1_pred_prior, plotfun = "scatter_avg") 
pp_check(rstanarm_1_pred_prior, plotfun = "error_hist", nreps = 6)

# BRM ++++++++++++++++++++++++

brm_1_pred <- brm(GRE.Score ~ CGPA, 
                  data=d, 
                  family="gaussian")
posterior_summary(brm_1_pred) 
print(stan_fit_1_pred, pars = c('b0', 'b1', 'sigma') )
plot(conditional_effects(brm_1_pred), points = TRUE, rug = TRUE)

# Priors ---------------------

prior_summary(brm_1_pred)

my_prior <- c(prior(student_t(3, 0, 2.5), class = "b"))
brm_1_pred_prior <- brm(GRE.Score ~ CGPA, 
                        data=d, 
                        family="gaussian",
                        prior=my_prior)
posterior_summary(brm_1_pred_prior)
prior_summary(brm_1_pred_prior)



```


# Multiple Regression

# Regression Practice
```{r}
d$TOEFL.Score_Stand <- as.numeric(scale(d$TOEFL.Score))
d$University.Rating_Stand <- as.numeric(scale(d$University.Rating))
d$SOP_Stand <- as.numeric(scale(d$SOP))
d$LOR_Stand <- as.numeric(scale(d$LOR)) 
d$CGPA_Stand <- as.numeric(scale(d$CGPA)) 
d$GRE_Stand <- as.numeric(scale(d$GRE.Score))

fit <- stan_glm(Chance.of.Admit ~ LOR + GRE.Score + CGPA + TOEFL.Score + Research + University.Rating + SOP, data = d)
summary(fit)

```
